Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-01 paper/paper.tex
dyck wip

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

@@

\paragraph{Regular Expressions} The theory of regular expressions emerges as a
subtheory of \theoryabbv by restricting the typing connectives to only use
$\otimes$ and $\oplus$ generated over $\bot$, $I$, and characters drawn from $\Sigma$.

@@

\subsection{LL(1)}
%% 4. LL(1) parser

\steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}

>>>>

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} by
Chomsky if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

@@

\paragraph{Regular Expressions} The theory of regular expressions emerges as the
subtheory of \theoryabbv generated by
$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
\newcommand{\balanced}{\mathsf{balanced}}
\newcommand{\Dyck}{\mathsf{Dyck}}
Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
cut to the essence of what it means to be context free.

$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
the empty string is balanced, and the $\balanced$ constrcutors builds a balanced
parse by wrapping an already balanced parse in an additional set of parentheses
then following it with another balanced parse.

\begin{figure}
\begin{align*}
\data &~\Dyck : L~\where\\
      & \nil : \ltonl {\Dyck} \\
      & \balanced : \ltonl {(\stringquote{(} \lto \Dyck \lto \stringquote{)} \lto \Dyck \lto \Dyck)}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:dyckinductive}
\end{figure}

\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
  handle 10's of thousands of characters in about 30s}

We construct a parser for $\Dyck$ by building a determinstic automaton $M$ such
that $\Trace_{M}(M.\init) \cong \Dyck$.

\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
approach to parse grammars like $\Dyck$ is to employ a stack to count how many
open parentheses have been seen so far. We may think of
$M$ as performing the same task as a stack machine, except instead of storing
the number of open parentheses in a distinguished stack of memory, $M$ operates
by storing this data in the current state. This is made feasible by specifying
$M$ almost identically to how we defined DFAs, except we omit the
finiteness condition on the states of the automaton.

\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }

% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
% from classical language theory, where the recognizer for $D$ would be encoded as a
% pushdown automaton with finitely many
% states and stack of memory. Such PDAs are encodable in our theory, although they
% are a little cumbersome to use. Moreover, they
% seem overfit to classical reasoning as they don't preserve the grammar up to
% strong equivalence, only weak equivalence in general. \steven{cite Lambek
%   automata and state that is is a secondary goal to preserve as much structure as possible}
% By moving the stack of memory (the nats) into the
% state, we obtain infinitely many states, which is fine bc the states need only
% be a type for us to interface with it.

\steven{TODO sketch proof of equivalence. Mention that it is mechanized}

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial, accepting] (0) {$0$};
    \node[state, right of=0] (1) {$1$};
    \node[state, right of=1] (2) {$2$};
    \node[right of=2] (3) {$\dots$};
    \node[state, below of=0] (fail) {$\mathsf{fail}$};

    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
              (1) edge[below, bend left] node{\stringquote{)}} (0)
              (1) edge[above, bend left] node{\stringquote{(}} (2)
              (2) edge[below, bend left] node{\stringquote{)}} (1)
              (2) edge[above, bend left] node{\stringquote{(}} (3)
              (3) edge[below, bend left] node{\stringquote{)}} (2)
              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
              (0) edge[left] node{\stringquote{)}} (fail);
  \end{tikzpicture}
  \caption{Automaton $M$ for the Dyck grammar}
  \label{fig:DyckAutomaton}
\end{figure}

% \subsection{LL(1)}
% %% 4. LL(1) parser

% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}
